\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}

\usepackage[margin=1in]{geometry}

\title{NASIC problem specification}
\author{Bryan O'Gorman}
\date{May 2016}

\begin{document}

\maketitle

\section{Introduction}
Notation:
\begin{itemize}
    \item $I$: num. airfields
    \item $J$: num. tasks
    \item $L$: num. aircraft types
    \item $b(i, l)$: num. $l$-type aircraft available at airfield $i$
    \item $B(i)$: max. num. aircraft able to be dispatched from airfield $i$
    \item $A(l)$: total num. $l$-type aircraft available
    \item $m(j, l)$: how much $l$-type aircraft contribute to task $j$
    \item $M(j)$: how much task $j$ needs
    \item $r(l, l', j)$: how much coverage $l'$-type aircraft provide for $l$-type aircraft executing task $j$
    \item $R(i, j, l)$: how much coverage each $l$-type aircraft from airfield $i$ executing task $j$ needs
    \item $C(i, j, l)$: Range of $l$-type aircraft from airfield $i$ executing task $l$
    \item $h(l)$: Range of $l$-type aircraft
    \item $N(i, j)$: set of airfields from which an aircraft can cover another aircraft executing task $j$ from airfield $i$  
    \item $Q(i, j, l, i')$: set of aircraft types in airfield $i'$ that can cover an $l$-type aircraft from airfield $i$ executing task $j$
\end{itemize}


\section{Constraints}

\begin{itemize}
\item 
\textbf{Limited cover availability:}
Only certain aircrafts from certain airfields are available for cover
\begin{equation} \label{eqn:constraint_cover_availability}
    y_{i, l, i', j', l'} \text{ only for } i \in N(i', j') \text{ and } l \in Q(i', j', l', i)
\end{equation}

\item
\textbf{Limited airfield and type resource:}
The sum of $l$-type primary and cover aircrafts dispatched from a airfield $i$ cannot exceed the number of $l$-type aircrafts stationed there
\begin{equation} \label{eqn:constraint_airfield_type_resource}
    \sum_j x_{i, j, l} + \sum_{i', j', l'}  y_{i, l, i', j', l'} \leq b(i, l) \qquad  \forall i, l 
\end{equation}

\item
\textbf{Limited airfield resource:}
The sum of any type of primary and cover aircrafts dispatched from a airfield $i$ cannot exceed the total number aircrafts stationed there
\begin{equation} \label{eqn:constraint_airfield_total_resource}
    \sum_{j, l} x_{i,j,l} + \sum_{l, i', j', l'} y_{i, l, i', j', l'} \leq B(i) \qquad \forall i
\end{equation}

\item
\textbf{Limited type resource:}
The sum of $l$-type primary and cover aircrafts dispatched from all airfield cannot exceed the total number of $l$-type aircrafts
\begin{equation} \label{eqn:constraint_type_total_resource}
    \sum_{i, j} x_{i, j, l} + \sum_{i, i', j', l'} y_{i, l, i', j', l'} \leq A(l) \qquad \forall l
\end{equation}

\item
\textbf{Tasks must be fulfilled:}
The sum of contributions of all aircrafts dispatched to fulfill task $j$ must exceed the resource needed by that task
\begin{equation}
    \sum_{i, l} m(j, l) x_{i, j, l} \geq M(j)  \qquad \forall j
\end{equation}

\item
\textbf{Aircraft range is limited:}
We can not dispatch aircrafts for tasks $j$ if the necessary range is beyond the maximum range of these aircrafts
\begin{equation}
    h(l) < C(i, j, l) \Rightarrow x_{i, j, l} = 0 \qquad \forall i, j, l 
\end{equation}

\item
\textbf{Cover must be provided:}
The sum of cover contributions to must exceed the cover needed.
\begin{equation}
    \sum_{i', l'} r(l, l', j) y_{i', l', i, j, l} \geq R(i, j, l) \qquad \forall i, j, l
\end{equation}
\end{itemize}

\section{Instance ensembles}
Simplifications:
\begin{itemize}
    \item Set $A(l) = \sum_{i} b(i, l)$
    \item Set $B(i) = \sum_{l} b(i, l)$
    \item Have $y_{i,l,i',j',l'}$ only when $(i, l) \in K_2(i', j', l')$
    \item Have $x_{i, j, l}$ only when $l \in K_1(i, j)$
    \item $r(l, l', j) = r(j) = 1$
    \item $R(i, j, l) = R(i, j) = 1$
\end{itemize}

\noindent
Oversimplifications:
\begin{itemize}
\item $L = 1$
\item $I$ small
\item $J$ small
\item $m(j, l) = 1$
\item $M(j)$ small 
\item $b(i, l) = b(i)$
\item (Maybe) $K_2(i', j', l') = I \times L = I \times \{1\}$
\item (Maybe) $K_1(i, j) = L$
\end{itemize}

\noindent
Remaining:
\begin{itemize}
\item $I$, $J$, $L$
\item $b(i, l)$
\item $m(j, l)$, $M(j)$
\item $r(l, l', j)$, $R(i, j, l)$
\item $N(i, j)$, $C(i, j, l)$, $h(l)$, $Q(i, j, l, i') \Rightarrow K_1, K_2$
\end{itemize}

\noindent 
Reduced problem:
$\mathbf x = \{x_{i, j}\}$, $\mathbf y = \{y_{i, i' ,j'}\}$
\begin{itemize}
\item $\forall i$: $\sum_j x_{i,j} + \sum_{i', j'} y_{i, i', j'} \leq b(i,l)$
\item $\forall i$: $\sum_j x_{i,j} + \sum_{i', j'} y_{i, i', j'} \leq B(i)$
\item \eqref{eqn:constraint_type_total_resource} implies \eqref{eqn:constraint_airfield_type_resource}
\item $\forall j$: $\sum_{i} x_{i, j} \geq M(j)$
\item $\forall i, j$: $\sum_{i'} y_{i', i, j} \geq 1$
\end{itemize}
(For all sums, $j \in [J]$, $i \in [I]$.)

\begin{itemize}
\item $I$ = 2, 3, 4
\item $J$ = $I$, $2I$, $3I$
\item $M(j)$ = 1, 2, 3
\item $b(i) = 1, 2, \ldots, 5$
\end{itemize}
Is this hard?

\section{QUBO}


\end{document}
